Procrastination and Dementia

### 2025-04-09 analysis/discrete-markov.qmd
Site re-render :collision:

pivot_and_factorise <- function(data) {
  # Converts cognitive function data from wide to long format and factorizes key variables.
  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
  # variables (Gender, Education, status) to factors with meaningful labels.
  # Arguments:
  #   - data: Dataset containing cognitive function variables in wide format
  # Returns:
  #   - Long-format dataset with factorized variables, ordered by ID and wave
  # 
  data |>
    tidyr::pivot_longer(
      cols = starts_with("cogfunction"), names_to = "wave",
      names_prefix = "cogfunction", values_to = "status") |>
    mutate(
      Gender = factor(Gender, levels = c(0, 1)),
      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
      wave = factor(wave),
      status = factor(status, 
                      levels = c("Normal Cognition", "MCI", "Dementia"),
                      labels = c(1, 2, 3))) |>
    relocate(wave, .after = ID) |>
    relocate(status, .after = wave)
}

@@

tidy_output <- function(fit) {
  output <- broom::tidy(fit, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
    mutate(
      term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Education_tri1" ~ "High school degree vs. No education",
      term == "Education_tri2" ~ "Further education vs. No education",
      term == "Total_dep_2016" ~ "Depression Scores (2016)",
      term == "Total_p" ~ "Procrastination",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      TRUE ~ term)) |>
  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))

  if(any(output$y.level == "1")) {
    output |>
      mutate(y.level = case_when(
      y.level == "1" ~ "MCI - NC",
      y.level == "3" ~ "MCI - Dementia"))
  } else {
    output |>
      mutate(y.level = case_when(
      y.level == "2" ~ "NC - MCI",
      y.level == "3" ~ "NC - Dementia"))
  }
}

@@

@@

```{r covariates}
cols <- c("ID", "Gender", "Age", "Education_tri", "Cardio_risk_16", "Total_dep_2016", "Total_p")
```

@@

```{r stack-data}
data_stack <- data_stack |>
  mutate(status_prev = lag(status), .after = status) |>
  ungroup() |> 
  filter(wave != 2016, !is.na(Total_p), Age >= 50)

@@

We evaluate model improvement using likelihood ratio tests:

@@

```{r model-comparison}
anova(fit_1a, fit_2a, fit_3a) |>
  knitr::kable(caption = "Nested Model Comparison") |>
  kableExtra::kable_styling(
    bootstrap_options = c("striped", "hover"),
    full_width = FALSE)
```

@@

Below we present the estimated parameters for the stationary models. We will transform the estimates to odds ratios and code them based on significance.

```{r model-output}
#| code-fold: true
#| code-summary: "Check out my code"
#| 
stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
  mutate(
    # Transforming to odds ratios
    estimate = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high),

    # Mutating to factors
    y.level = factor(
      y.level, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression Scores (2016)",
                 "Procrastination", "Previous state: MCI", "Previous state: Dementia")),

    # Creating a colour code
    positive = case_when(
      estimate > 1 & p.value < 0.05 ~ "Positive",
      estimate < 1 & p.value < 0.05 ~ "Negative",
      TRUE ~ "NS"),

    positive = factor(positive, levels = c("Positive", "Negative", "NS"))
    )

head(stationary_results)
```

@@

fig_4a <- stationary_results |>
  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
  ggplot(aes(x = estimate, y = y.level, colour = positive)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  ggstance::geom_pointrangeh(
    aes(xmin = conf.low, xmax = conf.high),
    position = ggstance::position_dodgev(height = 0.5),
    size = 1.25,
    fatten = 3) +
  scale_colour_manual(values = c(
    "Positive" = "#0072B2", 
    "Negative" = "#E69F00", 
    "NS"       = "#B2BEB5")) +
  labs(title = "Odds Ratios from Multinomial Model",
       x = "Odds Ratio", y = "Predictor") +
  guides(colour = "none") +
  facet_wrap(~ term, scales = "free_x") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())

@@

Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood). However, this time, we will also compare the best fitted non-stationary models `fit_3`.

```{r non-stationary-goodness}
#| code-fold: true
#| code-summary: "Check out my code"
#| layout-nrow: 2

@@

# Outputting as table
anova(fit_3a, fit_4a, fit_5a, fit_6a) |>
  mutate(
    Model = c("Stationary", "Additive", "State-Time Interactions", "Full Interactions"),
    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  cbind(comparison_table) |>
  DT::datatable(
    options = list(
      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    ),
    rownames = FALSE,
    caption = "Likelihood Ratio Tests"
  )
```

**Looks like adding in time as an additive term improved things**

@@

```{r non-stationary-output}
non_stationary_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))

head(non_stationary_results)
```

@@

Let's visualize the predicted transition probabilities from the non-stationary model @fig-non-stationary. We will show probability curves by procrastination level

```{r non-stationary-visualise}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 12
#| fig-height: 10
#| label: fig-non-stationary
#| fig-cap: "Predicted transition probabilities from the non-stationary model."
#| 
expand.grid(
  Gender = factor(0),
  Age = mean(data_stack$Age),
  Education_tri = factor(0),
  Total_dep_2016 = mean(data_stack$Total_dep_2016),
  status_prev = levels(data_stack$status_prev),
  Total_p = seq(0, 60, length = 200),
  wave = unique(data_stack$wave)) |>
  modelr::add_predictions(model = fit_4a, var = "pred", type = "probs") |>
  tidy_predictions() |>
  mutate(wave = factor(wave)) |>
  ggplot(aes(x = Total_p, y = prob, color = status_prev)) +
  geom_line(linewidth = 1) +
  ggokabeito::scale_color_okabe_ito() +
  labs(
    title = "Time-Varying Transition Probabilities",
    x = "Total Procrastination", 
    y = "Predicted Probability", color = "Previous State") +
  facet_grid(wave ~ status, labeller = labeller(
    wave = function(x) paste("Years since baseline:", x),
    status = function(x) paste("Transition to:", x)
  )) +
  theme(panel.spacing = unit(1, "lines"))

@@

@@

Let's visualize the estimated time-varying transition matrices @fig-non-stationary-matrix-plot

@@

@@

>>>>

pivot_and_factorise <- function(data) {
  # Converts cognitive function data from wide to long format and factorizes key variables.
  # Pivots multiple cogfunction columns into wave/status pairs and converts categorical
  # variables (Gender, Education, status) to factors with meaningful labels.
  # Arguments:
  #   - data: Dataset containing cognitive function variables in wide format
  # Returns:
  #   - Long-format dataset with factorized variables, ordered by ID and wave
  #

  # Pivoting depression variables
  # depression <- data |>
  #   select(starts_with("Total_dep")) |>
  #   tidyr::pivot_longer(
  #     cols = everything(), names_to = "Remove", values_to = "Depression") |>
  #   select(Depression)

  # Pivoting cardio variables
  # cardio <- data |>
  #   select(starts_with("Cardio_risk")) |>
  #   tidyr::pivot_longer(
  #     cols = everything(), names_to = "Remove", values_to = "Cardio") |>
  #   select(Cardio)

  data |>
    # select(!starts_with(c("Total_dep", "Cardio_risk"))) |>
    tidyr::pivot_longer(
      cols = starts_with("cogfunction"), names_to = "wave",
      names_prefix = "cogfunction", values_to = "status") |>
    mutate(
      Gender = factor(Gender, levels = c(0, 1)),
      Education_tri = factor(Education_tri, levels = c(0, 1, 2)),
      wave = factor(wave),
      status = factor(status, 
                      levels = c("Normal Cognition", "MCI", "Dementia"),
                      labels = c(1, 2, 3))) |>
    relocate(wave, .after = ID) |>
    relocate(status, .after = wave)
    #cbind(depression, cardio)
}

@@

tidy_output <- function(fit, multiplicative = FALSE) {
  output <- broom::tidy(fit, conf.int = TRUE) |>
    filter(term != "(Intercept)") |>
  mutate(across(c(estimate:p.value), ~ round(x = ., digits = 3)))

  if(any(output$y.level == "1")) {
    output <- output |>
      mutate(y.level = case_when(
      y.level == "1" ~ "MCI - NC",
      y.level == "3" ~ "MCI - Dementia"))
  } else {
    output <- output |>
      mutate(y.level = case_when(
      y.level == "2" ~ "NC - MCI",
      y.level == "3" ~ "NC - Dementia"))
  }

  if(multiplicative == FALSE) {
    output <- output |>
      mutate(
      term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Education_tri1" ~ "High school degree vs. No education",
      term == "Education_tri2" ~ "Further education vs. No education",
      term == "Total_dep_2016" ~ "Depression Scores (2016)",
      term == "Total_p" ~ "Procrastination (2020)",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      TRUE ~ term))
  } else {
    output <- output |>
      mutate(
      term = case_when(
      term == "Gender1" ~ "Being female",
      term == "Education_tri1" ~ "High school degree vs. No education",
      term == "Education_tri2" ~ "Further education vs. No education",
      term == "Depression" ~ "Depression Scores (2016)",
      term == "Total_p" ~ "Procrastination (2020)",
      term == "status_prev2" ~ "Previous state: MCI",
      term == "status_prev3" ~ "Previous state: Dementia",
      term == "wave" ~ "Time",
      term == "Gender1:wave" ~ "Gender & Time",
      term == "Age:wave" ~ "Age & Time",
      term == "Education_tri1:wave" ~ "Education (0 - 1) & Time",
      term == "Education_tri2:wave" ~ "Education (0 - 2) & Time",
      term == "Depression:wave" ~ "Depression & Time",
      term == "Total_p:wave" ~ "Procrastination & Time",
      term == "status_prev2:wave" ~ "Previous state: MCI & Time",
      term == "status_prev3:wave" ~ "Previous state: Dementia & Time",
      TRUE ~ term))
  }

  return(output)
}

@@

create_additive_predictions <- function(
    data, model, var, var_seq, hold_constant = list(Gender = factor(0), Education_tri = factor(0))) {

  # Set default prediction sequence based on variable type
  if (is.null(var_seq)) {
    if (is.numeric(data[[var]])) {
      predict_seq <- seq(min(data[[var]], max(data[[var]]), length = 200))
    } else if (is.factor(data[[var]])) {
      predict_seq <- levels(data[[var]])
    } else {
      predict_seq <- unique(data[[var]])
    }
  }

  # Create list of all variables needed for prediction
  all_vars <- all.vars(formula(model)[-2])  # Get RHS variables from model formula

  # Create base grid with all variables EXCEPT the prediction variable
  base_grid <- expand.grid(
    lapply(all_vars, function(v) {
      if (v == var) return(NULL)  # Skip prediction variable

      # Special cases - keep all levels
      if (v %in% c("status_prev", "wave")) {
        return(unique(data[[v]]))
      }

      if (v %in% names(hold_constant)) {
        # Use user-specified constant value
        hold_constant[[v]]
      } else if (is.numeric(data[[v]])) {
        # Use mean for numeric variables
        mean(data[[v]], na.rm = TRUE)
      } else if (is.factor(data[[v]])) {
        # Use first level for factors
        factor(levels(data[[v]])[1], levels = levels(data[[v]]))
      } else {
        # Default to most common value
        names(sort(table(data[[v]]), decreasing = TRUE))[1]
      }
    }) |> purrr::compact() |> setNames(all_vars[all_vars != var]),
    KEEP.OUT.ATTRS = FALSE
  )

  # Create full prediction grid
  pred_grid <- base_grid |> 
    tidyr::crossing(!!var := var_seq) |>
    modelr::add_predictions(model = model, var = "pred", type = "probs") |>
    tidy_predictions() |>
    mutate(wave = factor(wave))

  return(pred_grid)
}

plot_additive_predictions <- function(data, var, x_label, subtitle) {

  # Generating predictions
  data |>
    ggplot(aes(x = as.numeric(.data[[var]]), y = prob, color = status_prev)) +
    geom_line(linewidth = 1) +
    ggokabeito::scale_color_okabe_ito() +
    labs(
      title = "Transition Probabilities (Additive Model)",
      subtitle = subtitle,
      x = x_label, 
      y = "Predicted Probability", 
      color = "Previous State"
    ) +
    facet_grid(wave ~ status, labeller = labeller(
      wave = function(x) case_when(
        x == 2 ~ "Year = 2018",
        x == 3 ~ "Year = 2020",
        x == 4 ~ "Year = 2022",
      ),
      status = function(x) paste("Transition to:", x)
    )) +
    theme(
      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
      panel.spacing = unit(1, "lines")) 
}

@@

```{r covariates}
cols <- c(
  "ID", "Gender", "Age", "Education_tri",
  paste0("Cardio_risk_", seq(16, 22, by = 2)), 
  paste0("Total_dep_", seq(2016, 2022, by = 2)),
  "Total_p")
```

@@

```{r stack-data}
data_stack <- data_stack |>
  mutate(status_prev = lag(status), .after = status) |>
  filter(!is.na(Total_p), Age >= 50) |>
  ungroup() |> 
  filter(wave != 2016)

@@

We evaluate model improvement (@tbl-stationary-comparison) using likelihood ratio tests:

@@

```{r model-comparison}
#| label: tbl-stationary-comparison
#| tbl-cap: "Likelihood ratio test for stationary models"
#| tbl-cap-location: bottom

anova(fit_1a, fit_2a, fit_3a) |>
  mutate(
    Model = c("Baseline", "Procrastination", "Procrastination + Previous Status"),
    `Resid. Dev`= round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  DT::datatable(
    options = list(
      pageLength = 3,
      dom = "t",
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE
  )
```

@@

Below we present the estimated parameters for the stationary models (@tbl-stationary-output). We will transform the estimates to odds ratios and colour them based on significance.

```{r model-output}
#| code-fold: true
#| code-summary: "Check out my code"
#| label: tbl-stationary-output
#| tbl-cap: "Estimated parameters for the stationary model"
#| tbl-cap-location: bottom

stationary_results <- rbind(tidy_output(fit_3a), tidy_output(fit_3b)) |>
  rename(transition = y.level) |>
  mutate(
    # Transforming to odds ratios
    estimate = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high),

    # Mutating to factors
    transition = factor(
      transition, 
      levels = c("NC - MCI", "MCI - NC", 
                 "NC - Dementia", "MCI - Dementia")),
    term = factor(
      term, 
      levels = c("Being female", "Age", "High school degree vs. No education",
                 "Further education vs. No education", "Depression Scores (2016)",
                 "Procrastination (2020)", "Previous state: MCI", 
                 "Previous state: Dementia")),

    # Creating a colour code
    colour = case_when(
      estimate > 1 & p.value < 0.05 ~ "Positive",
      estimate < 1 & p.value < 0.05 ~ "Negative",
      TRUE ~ "NS"),

    colour = factor(colour, levels = c("Positive", "Negative", "NS"))
    )

stationary_results |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
  DT::datatable(
    options = list(
      pageLength = 6,
      dom = "tip",
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE
  )
```

@@

fig_4a <- stationary_results |>
  filter(!term %in% c("Previous state: MCI", "Previous state: Dementia")) |>
  ggplot(aes(x = estimate, y = transition, colour = colour)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "gray50") +
  ggstance::geom_pointrangeh(
    aes(xmin = conf.low, xmax = conf.high),
    position = ggstance::position_dodgev(height = 0.5),
    size = 1.25,
    fatten = 3) +
  scale_colour_manual(values = c(
    "Positive" = "#0072B2", 
    "Negative" = "#E69F00", 
    "NS"       = "#B2BEB5")) +
  labs(title = "Odds ratios (stationary model)",
       x = "Odds Ratio", y = "Predictor") +
  guides(colour = "none") +
  facet_wrap(~ term, scales = "free_x") +
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    axis.title = element_text(size = 12, face = "bold"),
    strip.text = element_text(size = 10, face = "bold"),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())

@@

Again, we will evaluate model improvement using likelihood ratio tests (@eq-liklihood; @tbl-non-stationary-comparison). However, this time, we will also compare the best fitted non-stationary models `fit_3`.

```{r non-stationary-goodness}
#| code-fold: true
#| code-summary: "Check out my code"
#| label: tbl-non-stationary-comparison
#| tbl-cap: "Likelihood ratio test for non-stationary models"
#| tbl-cap-location: bottom

@@

# Outputting as table
anova(fit_3a, fit_4a, fit_5a, fit_6a) |>
  mutate(
    Model = c("Stationary", "Additive", "State-Time Interactions", "Full Interactions"),
    `Resid. Dev` = round(`Resid. Dev`, digits = 3),
    `LR stat.` = round(`LR stat.`, digits = 3),
    `Pr(Chi)` = round(`Pr(Chi)`, digits = 3)
    ) |>
  cbind(comparison_table) |>
  DT::datatable(
    options = list(
      pagelength = 4,
      dom = "t",
      ordering = FALSE,
      columnDefs = list(list(className = 'dt-center', targets = "_all"))
    ),
    rownames = FALSE
  )
```

**This is interesting!! Looks like adding in time as an additive term improved things. However, having full time interactions is better than having state-time interactions**

::: {.panel-tabset .nav-pills}
## Additive Model

@@

```{r non-stationary-output}
additive_results <- rbind(tidy_output(fit_4a), tidy_output(fit_4b))

additive_results |>
  rename(transiton = y.level) |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
  DT::datatable(
    options = list(
      pageLength = 6,
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE,
    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
                 "Statistic", "p-val", "Lower CI", "Upper CI")
  )
```

@@

Let's visualize the predicted transition probabilities from the non-stationary model.

- @fig-non-stationary-age shows the predicted transition probabilities for age
- @fig-non-stationary-depression shows the predicted transition probabilities for depression
- @fig-non-stationary-education shows the predicted transition probabilities for per education level
- @fig-non-stationary-procrastination shows the predicted transition probabilities for procrastination

::: {.panel-tabset .nav-pills}

## Age

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-age
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for age)."
#|

create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
  plot_additive_predictions(var = "Age", x_label = "Age", subtitle = "By Age")

@@

## Depression

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-depression
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for depression)."
#|

create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_dep_2016", var_seq = seq(0, 8, by = 1)) |>
  plot_additive_predictions(var = "Total_dep_2016", x_label = "Depression", subtitle = "By depression")

## Education Level

```{r}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-education
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for having a high school degree)."

create_additive_predictions(data = data_stack, model = fit_4a, var = "Education_tri", var_seq = factor(c(0:2)), hold_constant = list(Gender = factor(0))) |>
  plot_additive_predictions(var = "Education_tri", x_label = "Education", subtitle = "By education level") +
  geom_point(size = 2) +
  scale_x_continuous(breaks = c(0, 1, 2), labels = c("No Education", "High School", "Further Education")) +
  theme(axis.text.x = element_text(size = 8, angle = 60, vjust = 0.90, hjust = 1))
```

## Procrastination

```{r non-stationary-visualise-procrastination}
#| code-fold: true
#| code-summary: "Check our my code"
#| fig-width: 10
#| fig-height: 8
#| label: fig-non-stationary-procrastination
#| fig-cap: "Predicted transition probabilities from the non-stationary model (for procrastination)."
#|

create_additive_predictions(data = data_stack, model = fit_4a, var = "Total_p", var_seq = seq(0, 60, by = 1)) |>
  plot_additive_predictions(var = "Total_p", x_label = "Procrastination", subtitle = "By procrastination")

```
:::

@@

@fig-non-stationary-matrix-plot shows the estimated time-varying transition matrices

@@

## Multiplicative Model

Key coefficients from the selected non-stationary model `fit_6`

```{r multiplicative-output}
multiplicative_output <- rbind(
  tidy_output(fit_6a, multiplicative = TRUE),
  tidy_output(fit_6b, multiplicative = TRUE))

multiplicative_output |>
  mutate(across(c(estimate, conf.low, conf.high), ~ round(x = ., digits = 3))) |>
  DT::datatable(
    options = list(
      pageLength = 6,
      ordering = FALSE,
      columnDefs = list(list(className = "dt-center", targets = "_all"))
    ),
    rownames = FALSE,
    colnames = c("Transition", "Predictor", "Estimate", "Standard Error",
                 "Statistic", "p-val", "Lower CI", "Upper CI")
  )

```
:::

@@

```{r test-code}
#| include: false
#| eval: false
preds <- create_additive_predictions(data = data_stack, model = fit_4a, var = "Age", var_seq = seq(50, 97, by = 5)) |>
  # Add color column based on transition significance
  mutate(
    transition_type = paste(status_prev, status, sep = " → "),
    colour = case_when(
      transition_type == "Normal Cognition → Normal Cognition" ~ "#E69F00",
      transition_type == "Normal Cognition → MCI" ~ "#E69F00",
      transition_type == "Normal Cognition → Dementia" ~ "#B2BEB5",
      transition_type == "MCI → Normal Cognition" ~ "#56B4E9",
      transition_type == "MCI → MCI" ~ "#56B4E9",
      transition_type == "MCI → Dementia" ~ "darkgrey",
      transition_type == "Dementia → Normal Cognition" ~ "#009E73",
      transition_type == "Dementia → MCI" ~ "#009E73",
      transition_type == "Dementia → Dementia" ~ "#009E73"
    ))

ggplot(preds, aes(Age, prob, color = colour, group = interaction(wave, transition_type))) +
  geom_line(linewidth = 1) +
  scale_color_identity("Previous State", 
                      breaks = c("#E69F00", "#56B4E9", "#009E73"),
                      labels = c("Normal Cognition", "MCI", "Dementia"),
                      guide = "legend") +
  labs(
    title = "Transition Probabilities (Additive Model)",
    subtitle = "By Age",
    x = "Age", 
    y = "Predicted Probability",
    caption = "*Note:* Grey lines indicate non-significant transitions",
    ) +
    facet_grid(wave ~ status, labeller = labeller(
      wave = function(x) case_when(
        x == 2 ~ "Year = 2018",
        x == 3 ~ "Year = 2020",
        x == 4 ~ "Year = 2022",
      ),
      status = function(x) paste("Transition to:", x)
    )) +
    theme(
      plot.subtitle = element_text(hjust = 0.5, size = 12, face = "bold"),
      plot.caption = ggtext::element_markdown(),
      panel.spacing = unit(1, "lines"))

### 2025-04-09 manuscript/manuscript.qmd
Site re-render :collision:

# Method {#method .unnumbered}

@@

For this study, we focused on four waves of HRS data (from 2016 to 2022). Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave. These experimental modules, administered at the end of the core HRS interview, consist of concise questionnaires designed to explore new topics or supplement existing core survey data. Each respondent receives only one experimental module, with sample sizes for each module constituting approximately 10% of the core sample. As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents. We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$ and those under 50 years of age $(n = 39)$. This resulted in a final analytic sample of $884$ respondents.

## Measures {#measures .unnumbered}

### Outcome: Cognitive Function and Cognitive Category {#outcome-cognitive-function-and-cognitive-category .unnumbered}

@@

### Predictor: Procrastination {#predictor-procrastination .unnumbered}

@@

### Confounders {#confounders .unnumbered}

@@

>>>>

# Method

@@

For this study, we focused on four biennial waves of HRS data (from 2016 to 2022). Specifically our study sample consisted of respondents who participated in an experimental module assessing procrastination during the 2020 wave. These experimental modules, administered at the end of the core HRS interview, consist of concise questionnaires designed to explore new topics or supplement existing core survey data. Each respondent receives only one experimental module, with sample sizes for each module constituting approximately 10% of the core sample. As a result, while the core HRS sample includes approximately $18,000$ respondents, our initial sample of interest consisted of $1,368$ respondents. We excluded respondents with missing cognitive assessment data for any wave $(n = 445)$, missing procrastination data $(n = 4)$, and those under 50 years of age $(n = 16)$. This resulted in a final analytic sample of $903$ respondents.

## Measures

### Outcome: Cognitive Function and Cognitive Category

@@

### Predictor: Procrastination

@@

### Confounders

@@

## Data analysis

All data analysis was carried out in R [@R2024]. Within the final dataset, $0.69\%$ of the data was missing, with all all missing values occurring within the educational attainment variable. To examine longitudinal patterns of cognitive transitions, we implemented a discrete time, first-order Markov model [@braun2020]. This framework estimates probabilities of transitioning between three discrete states: normal cognition (State 1), mild cognitive impairment (MCI; State 2), and dementia (State 3). The model relies on the Markov property, which assumes that the probability of transitioning to a future state depends only on the current state, not the full history of prior states.

### Model specification
Formally, let $X_t$ denote an individual’s cognitive state at time $t$, where $X_t \in S = \{1, 2, 3\}$. Under the Markov property, the probability of transitioning to state $j$ at time $t + 1$ depends only on state $i$ at time $t$, independent of all earlier states. We can write this conditional probability as:

$$
P(X_{t+1} = j \vert X_t = i, X_{t-1} = i_{t-1}, \dots X_0 = i_0) = P(X_{t+1} = j \vert X_t = i)
$$

These conditional probabilities, denoted $p_{ij} = P(X_{t+1} = j \vert X_t = i$ are called transition probabilities and can be organised into a $3 \times 3$ transition matrix $P$:

$$
P = 
\begin{bmatrix} 
p_{11} & p_{12} & p_{13} \\
p_{21} & p_{22} & p_{23} \\
p_{31} & p_{32} & p_{33} \\
\end{bmatrix}
$$

wherein, each row represents the cognitive state at time $t$, each column represents the cognitive state at time $t + 1$, and each entry $p_{ij}$ represents the probability of transitioning from state $i$ to state $j$.

### Non-stationary transition probabilities

In a stationary Markov model, transition probabilities $p_{ij}$ are assumed constant over time. However, cognitive transitions may vary with age, time since diagnosis, or other covariates. To account for this, we generalize to a non-stationary Markov model, where $p_{ij}^{(t)}$ depends on covariates $\boldsymbol{x}^{(t)}$

$$
p_{ij}^{(t)} = P(X_{t+1} = j \vert X_t = i, \boldsymbol{x}^{(t)})
$$

#### Multinomial Logistic Regression

We model these time-varying transition probabilities using multinomial logistic regression. For a system with $K = 3$ states (using state $K$ as reference), the log-odds of transitioning to state $j$ relative to state $K$ are linear functions of covariates $\boldsymbol{x}(t)$:

$$
log \left( \frac{p_{ij}^{(t)}}{p_{ik}^{(t)}} \right) = \alpha_j^{(t)} + \beta_j^{T \;(t)} \boldsymbol{x}^{(t)} \qquad \text{for } j = 1, \dots K-1
$$
where, $\alpha_j^{(t)}$ is the intercept for state $j$ at time $t$, $\beta_j^{(t)}$ is the vector of regression coefficients for state $j$ at time $t$, and $\boldsymbol{x}^{(t)}$ is the vector of covariates at time $t$. The model estimates the transition probabilities as a function of covariates, allowing us to assess how procrastination and other factors influence cognitive transitions.

# Results

In our sample of $903$ older American adults aged 50 years or older, $63.01\%$ were female $(n = 569)$, with a mean sample age of $67.73$ years $(\pm 9.84)$. At baseline, $84.81\%$ of the sample was classified as having normal cognition, $14.96\%$ as having MCI, and $1.35\%$ as having dementia.

The mean procrastination score was $36.73$ $(\pm 8.23)$, with a range of $12 - 60$. The sample characteristics are presented in Table \ref{tab:sample-characteristics}.
